---
title: "README"
author: "Stephen Lauer"
date: "`r Sys.Date()`"
output: md_document
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r library, include=FALSE}
library(dplyr)
library(tidyr)
library(kableExtra)

game_dat <- read.csv("data-raw/Octagon.csv")

game_stats <- game_dat %>%
    transmute(Date=lubridate::ymd(paste0("2021/", Date)),
              Team=Home,
              Opp=Away,
              Place="Home",
              GF=Home.Score,
              GA=Away.Score) %>%
    bind_rows(game_dat %>%
                  transmute(Date=lubridate::ymd(paste0("2021/", Date)),
                            Team=Away,
                            Opp=Home,
                            Place="Away",
                            GF=Away.Score,
                            GA=Home.Score)) %>%
    group_by(Opp) %>%
    mutate(GD=GF-GA,
           Opp_GFpg=(sum(GA) - GA)/(n() - 1),
           Opp_GApg=(sum(GF) - GF)/(n() - 1),
           Opp_GDpg=Opp_GFpg-Opp_GApg) %>%
    ungroup() %>%
    mutate(GF_above=GF-Opp_GApg,
           GA_below=Opp_GFpg-GA,
           G_better=GF_above + GA_below,
           Points=ifelse(GF > GA, 3,
                         ifelse(GF==GA, 1, 0)))

GP <- game_stats %>% filter(Team=="United States") %>% nrow()

```

## USMNT Octagon

CONCACAF has entered the Octagon, and I wanted to see which teams have had the hardest/easiest schedules so far and who has performed best relative to expectations.
What follows is super basic (you can find my exploratory analysis code in `inst/octagon.R`).
**I would love to do this analysis on xG instead of observed goals, please let me know if you have that data for all the games!**

* [Who Played Whom and Results](#who-played-whom-and-results)
* [Strength of Schedule](#strength-of-schedule)
* [Real Basic Result-Only Rankings](#real-basic-result-only-rankings)
* [Big Ugly Table](#big-ugly-table)

### Who Played Whom and Results

With 8 teams each playing `r GP` games so far, it's hard to understand everything that has happened so far looking at the traditional schedule/results format.
Sometimes you just want to see who everyone has played and the scores.
That's what we've got here, with each game represented twice, with both the home team and away team in the `Team` column:

```{r game-results, message=FALSE, warning=FALSE}
game_stats %>% 
    arrange(Team, Date) %>% 
    select(Date, Team, Opponent=Opp, Place, GF, GA, GD, Points) %>% 
    kable(format="html", booktabs=T, align="llllcccc") %>% 
    kable_styling() %>% 
    row_spec(seq(GP, nrow(game_stats), GP), extra_css="border-bottom: 1px solid")
```

## Strength of Schedule

To figure out who has played the hardest slate so far, I look at how a team's opponents fared in their other games and how the team then performed relative to that.
This does not yet account for number of home/away matches by each nation.
Here you can see each team and their opponents average goals for and against in other games:

```{r team-summaries, message=FALSE, warning=FALSE}
game_stats %>%
    group_by(Team) %>%
    summarize(Opp_GF=mean(Opp_GFpg),
              Opp_GA=mean(Opp_GApg),
              Opp_GD=Opp_GF-Opp_GA,
              Points=sum(Points)) %>% 
    arrange(desc(Opp_GD)) %>% 
    rename(`Opp Avg GF`=Opp_GF,
           `Opp Avg GA`=Opp_GA,
           `Opp Avg GD`=Opp_GD) %>% 
    kable(format="html", digits=2, booktabs=T, align="lcccc") %>% 
    kable_styling()
```

Panama has played the hardest schedule with only Honduras remaining before the second set of games.
The US has the easiest schedule as they have not yet played Mexico.

## Real Basic Result-Only Rankings

```{r team-summaries2, message=FALSE, warning=FALSE}
game_stats %>%
    group_by(Team) %>%
    summarize(Opp_GF=mean(Opp_GFpg),
              Opp_GA=mean(Opp_GApg),
              Opp_GD=Opp_GF-Opp_GA,
              GF_above=mean(GF_above),
              GA_below=mean(GA_below),
              G_better=GF_above + GA_below,
              Points=sum(Points)) %>% 
    arrange(desc(G_better), desc(Opp_GD)) %>% 
    rename(`Opp Avg GF`=Opp_GF,
           `Opp Avg GA`=Opp_GA,
           `Opp Avg GD`=Opp_GD,
           `GF above Opp Avg`=GF_above,
           `GA below Opp Avg`=GA_below,
           `GD above Avg`=G_better) %>% 
    kable(format="html", digits=2, booktabs=T, align="lccccccc") %>% 
    kable_styling()
```

Canada and Mexico are virtually tied at the top of the table after Canada's big win over Panama, which was the largest over expectation of any match to date.
El Salvador has plummeted from top 4 to bottom 2 in the last three games.

### Big Ugly Table

Here's a big ugly table with all of the stats for all of the games, if you're interested:

```{r all-results, message=FALSE, warning=FALSE}
game_stats <- game_dat %>%
    transmute(Date=lubridate::ymd(paste0("2021/", Date)),
              Team=Home,
              Opp=Away,
              Place="Home",
              GF=Home.Score,
              GA=Away.Score) %>%
    bind_rows(game_dat %>%
                  transmute(Date=lubridate::ymd(paste0("2021/", Date)),
                            Team=Away,
                            Opp=Home,
                            Place="Away",
                            GF=Away.Score,
                            GA=Home.Score)) %>%
    group_by(Opp) %>%
    mutate(GD=GF-GA,
           Opp_GFpg=(sum(GA) - GA)/(n() - 1),
           Opp_GApg=(sum(GF) - GF)/(n() - 1),
           Opp_GDpg=Opp_GFpg-Opp_GApg) %>%
    ungroup() %>%
    mutate(GF_above=GF-Opp_GApg,
           GA_below=Opp_GFpg-GA,
           G_better=GF_above + GA_below,
           Points=ifelse(GF > GA, 3,
                         ifelse(GF==GA, 1, 0)))

game_stats %>% 
    arrange(Team, Date) %>% 
    rename(Opponent=Opp,
           `Opp Avg GF`=Opp_GFpg,
           `Opp Avg GA`=Opp_GApg,
           `Opp Avg GD`=Opp_GDpg,
           `GF above Opp Avg`=GF_above,
           `GA below Opp Avg`=GA_below,
           `GD above Avg`=G_better) %>%
    kable(format="html", digits=2, booktabs=T, align="llllcccccccc") %>% 
    kable_styling() %>% 
    row_spec(seq(GP, nrow(game_stats), GP), extra_css="border-bottom: 1px solid")
```

I'll try to make these all sortable in the future.
